EBASE manuscript

### 2023-03-21 manu-draft.qmd
starting rgn revisions
more rgn edits

institute:
  - TBEP: Tampa Bay Estuary Program, St. Petersburg, Florida 33701 USA
  - PSU: Pennsylvania State University, University Park, Pennsylvania 16802 USA

@@

$$
\frac{\partial{C_d}}{\partial{t}} = [\,aPAR]\, - [\,r]\, - \frac{1}{H}\left[-bU_{10}^2\left(\frac{s_c}{600} \right)^{-0.5} \left(C_{Sat} - C_d \right )\right]
$$ {#eq-ebase}

where the terms in the square brackets are production (P), respiration (R), and gas exchange (D), respectively, modeled as a function of the change in dissolved oxygen ($\partial{C_d}$) per unit time ($\partial{t}$) of the input data. Production is modeled linearly as a function of surface irradiance ($PAR$, Watts m$^{-2}$) and $a$ (light efficiency, $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$). Respiration is modeled with the single parameter $r$ ($\text{mmol}/\text{m}^3/\text{d}$). Gas exchange is modeled using the formulation in @Wanninkhof14 as a wind-based parameterization for the gas transfer velocity [@Ho16] multiplied by the oxygen concentration gradient as the difference between the concentration measured in the surface water layer, $C_d$ (mmol m$^{-3}$), and the saturation (or equilibrium) concentration, $C_{Sat}$ (mmol m$^{-3}$), calculated from salinity (psu) and temperature ($^\circ$C) [@Millero81; @Garcia92]. The gas transfer velocity is modeled using wind speed squared at 10 meters above the water ($U_{10}^2$, m$^2$ s$^{-2}$), water column depth ($H$ in meters, as a fixed parameter or time series from a pressure sensor), and the Schmidt number ($S_c$) defined as the ratio of the kinematic viscosity of water to the molecular diffusivity of the gas for oxygen [from water temperature and salinity using the polynomial fit in @Wanninkhof14]. The $b$ parameter is estimated from the data and has units of $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$. As such, the three parameters estimated by EBASE are $a$, $r$, and $b$ using the required input data that includes time series of dissolved oxygen, water temperature, salinity, PAR, wind speed, and water column depth (`r run_reference(id = "ebaseiotab")`).

@@

$$
\frac{\partial{C_d}}{\partial{t}} = [\,aPAR^p]\, - [\,r\Theta^{T - \bar{T}}]\, - \left[-K\cdot1.0241^{T-\bar{T}}\left(C_{sat} - C_d \right )\right]
$$ {#eq-base}

where all parameters are defined as in @eq-ebase, with the additional parameters $p$, $\Theta$, and $K$ (the rearation coefficient). The BASE approach provides options to estimate $p$ and $\Theta$ from the data or to set as fixed values. A temperature dependency, $T$, is also included where respiration and gas exchange vary with temperature. The changes to @eq-base for EBASE were necessary for estuarine application and included 1) a linear relationship of production to PAR ($aPAR$) by setting $p = 1$; 2) constant respiration throughout the time series by removing the temperature dependency; and 3) use of a gas exchange parameterization commonly used in coastal or oceanic settings [i.e., @Grace15 used a littoral parameterization]. The justification for these changes is as follows.

@@

A common theory behind most gas exchange parameterizations for metabolism estimates is that the transfer of gases of relatively low solubility, like oxygen, is proportional to the concentration gradient in a thin film of water at the air--water interface. It is also assumed that the water at the top of the film is saturated with respect to the atmosphere, so that the concentration gradient across the thin film can be approximated as the difference between oxygen concentration measured in the surface water layer, $C_d$, and the saturation (or equilibrium) concentration $C_{Sat}$, which is usually calculated as a function of water temperature and salinity (the negative sign is needed because positive flux is set to be from water to air):

$$
D = -k_w \left(C_{sat}-C_d \right).
$$ {#eq-gasex}

@@

where $\bar{T}$ is the average water temperature over each 24-hour period in the diel oxygen time series and $K$ is the reaeration coefficient estimated from the data. @eq-gasexbase indicates that $k_w$ increases with temperature and depth. The temperature dependence is consistent with the known decrease of $k_w$ with the Schmidt number, but the depth dependence is difficult to justify. An expectation is that $k_w$ would increase as water depth decreases for a given current speed because turbulence generated at the sediment--water interface by currents would more easily reach the surface [@Ho16]. Based on these differences, the wind-based parameterization for gas exchange was used for EBASE [@Wanninkhof14]. Additionally, $C_{sat}$ was calculated using only temperature and salinity, removing the need for barometric pressure used in BASE.

@@

The metabolism estimates and their parameters returned by the Bayesian routine implemented in JAGS are affected by the prior distributions assigned to each. As for the BASE model, relatively uninformed prior distributions following a normal Gaussian distribution are used by default, although the priors can be changed based on previous knowledge of parameters specific to an ecosystem or as informed by other metabolic modeling approaches. The default priors were based on those that returned reasonable parameter estimates using a known time series created with forward-feeding oxygen mass balance inputs to @eq-ebase. [Fwoxy R package, @Arriola22]. All of the physical parameters were held constant, i.e., wind speed, temperature, and salinity. The other required inputs were the $a$, $r$, and $b$ parameters, chosen as 0.2 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$, 20 $\text{mmol}/\text{m}^3/\text{d}$, and 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$, respectively. An initial oxygen concentration was provided as 250 mmol/m$^3$ and the time series was created by forward addition of the inputs. As such, the dissolved oxygen output as change per unit time was modelled by EBASE using the constant the same physical parameters as input. Reasonable prior distributions by EBASE were identified (Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior]) based on the ability to return the known parameters:

$$
a \sim N(0, 1)T(0,)
$$ {#eq-aprior}

$$
r \sim N(0, 50)T(0,)
$$ {#eq-rprior}

$$
b \sim N(0, 0.1)T(0,0.502)
$$ {#eq-bprior}

Each follows a normal distribution with means and standard deviation defined in the parentheses (@fig-priorplot). The $T$ notation is JAGS model syntax indicating the distributions are truncated to a minimum (left) and maximum (right) value defined in the parentheses. All prior distributions were constrained to positive values based on known physical constraints and requirements of the model formula in @eq-ebase. Additionally, the prior distribution for $b$ was constrained to an upper limit of 0.504 (\~twice the default mean). @Wanninkhof14 states that the $b$ parameter typically does not vary by more than 20% and the standard deviation of the default prior is sufficiently high to expand beyond this range. Initial development of EBASE also showed that using an undefined upper limit of $b$ led to unstable and unreasonable parameter estimates.

>>>>

institute:
  - TBEP: Tampa Bay Estuary Program, St. Petersburg, Florida 33701 USA
  - PSU: The Pennsylvania State University, University Park, Pennsylvania 16802 USA

@@

$$
H\frac{dC}{dt} = aPAR - r - \left[bU_{10}^2\left(\frac{Sc}{600} \right)^{-0.5} \left(C_{Sat} - C \right )\right]
$$ {#eq-ebase}

where the terms in the square brackets are production (P), respiration (R), and gas exchange (D), respectively, modeled as a function of the change in dissolved oxygen ($dC$) per unit time ($dt$) of the input data. Production is modeled linearly as a function of surface irradiance ($PAR$, Watts m$^{-2}$) and $a$ (light efficiency, $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$). Respiration is modeled with the single parameter $r$ ($\text{mmol}/\text{m}^3/\text{d}$). Gas exchange is modeled using the formulation in @Wanninkhof14 as a wind-based parameterization for the gas transfer velocity [@Ho16] multiplied by the oxygen concentration gradient as the difference between the concentration measured in the surface water layer, $C$ (mmol m$^{-3}$), and the saturation (or equilibrium) concentration, $C_{Sat}$ (mmol m$^{-3}$), calculated from salinity (psu) and temperature ($^\circ$C) [@Millero81; @Garcia92]. The gas transfer velocity is modeled using wind speed squared at 10 meters above the water ($U_{10}^2$, m$^2$ s$^{-2}$), water column depth ($H$ in meters, as a fixed parameter or time series from a pressure sensor), and the Schmidt number ($Sc$) defined as the ratio of the kinematic viscosity of water to the molecular diffusivity of the gas for oxygen [from water temperature and salinity using the polynomial fit in @Wanninkhof14]. The $b$ parameter is estimated from the data and has units of $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$. As such, the three parameters estimated by EBASE are $a$, $r$, and $b$ using the required input data that includes time series of dissolved oxygen, water temperature, salinity, PAR, wind speed, and water column depth (`r run_reference(id = "ebaseiotab")`).

@@

$$
\frac{dC}{dt} = aPAR^p - r\Theta^{T - \bar{T}} + \left[K\cdot1.0241^{T-\bar{T}}\left(C_{sat} - C \right )\right]
$$ {#eq-base}

where all parameters are defined as in @eq-ebase, with the additional parameters $p$, $\Theta$, and $K$ (the rearation coefficient). The BASE approach provides options to estimate $p$ and $\Theta$ from the data or to set as fixed values. A temperature dependency, $T$, is also included where respiration and gas exchange vary with temperature. The changes to @eq-base for EBASE were necessary for estuarine application and included 1) a linear relationship of production to PAR ($aPAR$) by setting $p = 1$; 2) constant respiration by removing the temperature dependency; and 3) use of a gas exchange parameterization commonly used in coastal or oceanic settings that depends on wind speed and temperature [i.e., @Grace15 used a littoral parameterization]. The justification for these changes is as follows.

@@

A common theory behind most gas exchange parameterizations for metabolism estimates is that the transfer of gases of relatively low solubility, like oxygen, is proportional to the concentration gradient in a thin film of water at the air--water interface. It is also assumed that the water at the top of the film is saturated with respect to the atmosphere, so that the concentration difference across the thin film can be approximated as the difference between oxygen concentration measured in the surface water layer, $C$, and the saturation (or equilibrium) concentration $C_{Sat}$, which is usually calculated as a function of water temperature and salinity (the negative sign is needed because positive flux is set to be from water to air):

$$
D = -k_w \left(C_{sat}-C \right).
$$ {#eq-gasex}

@@

where $\bar{T}$ is the average water temperature over each 24-hour period in the oxygen time series and $K$ is the reaeration coefficient estimated from the data. @eq-gasexbase indicates that $k_w$ increases with temperature and depth. The temperature dependence is consistent with the known decrease of $k_w$ with the Schmidt number, but the depth dependence is difficult to justify. An expectation is that $k_w$ would increase as water depth decreases for a given current speed because turbulence generated at the sediment--water interface by currents would more easily reach the surface [@Ho16]. Based on these differences, the wind-based parameterization for gas exchange was used for EBASE [@Wanninkhof14]. Additionally, $C_{sat}$ was calculated using only temperature and salinity, removing the need for barometric pressure used in BASE.

@@

The metabolism estimates and their parameters returned by the Bayesian routine implemented in JAGS are affected by the prior distributions assigned to each. As for the BASE model, relatively uninformed prior distributions following a normal Gaussian distribution are used by default, although the priors can be changed based on previous knowledge of parameters specific to an ecosystem or as informed by other metabolic modeling approaches. Reasonable uninformed prior distributions for EBASE were chosen based on normal distributions with mean values based on approximate estimates from the literature [@Wanninkhof14;@Grace15] and standard deviations that were sufficiently large to allow the Bayesian routine to search an unconstrained parameter space.  The default priors for $a$, $r$, and $b$ are:

$$
a \sim N(0.2, 1)T(0,\infty) \;\; \left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)
$$ {#eq-aprior}

$$
r \sim N(20, 50)T(0,\infty) \;\; \text{mmol}/\text{m}^3/\text{d}
$$ {#eq-rprior}

$$
b \sim N(0.251, 0.1)T(0,0.502) \;\; \left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)
$$ {#eq-bprior}

where $N(\alpha, \beta)$ indicates a normal distribution with mean $\alpha$ and standard deviation $\beta$ and $T(\gamma, \delta)$ is 1 between $\gamma$ and $\theta$ and 0 elsewhere; $T$ has the effect of truncating the normal distribution to minimum and maximum values. All prior distributions were constrained to positive values based on known physical constraints and requirements of the model formula in @eq-ebase. Additionally, the prior distribution for $b$ was constrained to an upper limit of 0.502 (\~twice the default mean). Initial development of EBASE showed that using an undefined upper limit of $b$ led to unstable and unreasonable parameter estimates.  Examples of the prior distributions in Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior] are shown in @fig-priorplot.
